---
title: "Darwin Core mapping"
subtitle: "For: alien-mollusca-checklist"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
#  pdf_document:
#    df_print: kable
#    number_sections: yes
#    toc: yes
#    toc_depth: 3
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(readxl)         # To read Excel files
library(digest)         # To generate hashes
library(rgbif)          # To use GBIF services
library(googlesheets)   # To import and read Google spreadsheets 
```

# Read source data

Set column types:

```{r}
col_types = cols(
   .default = col_character(),
    other_col = col_double())
```

The original spreadsheet can be found [here](https://docs.google.com/spreadsheets/d/1hSXd6SZQUvI8uGbE_5-vmN8ZZz739D46HRt2aTFcb-s/edit#gid=0). We need to retrieve this spreadsheet and select the specific worksheet first:

Retrieve the spreadsheet:

```{r connect_google_spreadsheets}
retrieve_spreadsheet <- gs_title("alien mollusca checklist")
```

Select the data in the worksheet `checklist`:

```{r read_source_data}
input_data <- retrieve_spreadsheet %>% gs_read("checklist", col_types = col_types)
```

We want to add a copy of the source data to the repository:

```{r}
write_csv(input_data, here("data", "raw", "alien_mollusca_checklist_dump.csv"), na = "")
```

Preview data: 

```{r}
input_data %>% head()
```

# Process source data

## Tidy data

Clean data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %<>% 
  clean_names()
```

## Scientific names

Use the [GBIF nameparser](https://www.gbif.org/tools/name-parser) to retrieve nomenclatural information for the scientific names in the checklist:

```{r}
parsed_names <- input_data %>%
  distinct(scientific_name) %>%
  pull() %>% # Create vector from dataframe
  parsenames() # An rgbif function
```

All scientific names are of `type = SCIENTIFIC` and need no further processing.

## Taxon ranks

The nameparser function also provides information about the rank of the taxon (in `rankmarker`). Here we join this information with our checklist. Cleaning these ranks will done in the Taxon Core mapping:

```{r}
input_data %<>% left_join(
  select(parsed_names, scientificname, rankmarker),
  by = c("scientific_name" = "scientificname"))
```

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`.
Although information in `r_id` could be used as a `taxonID` here, we will not consider them as such as they are unstable (e.g. `unknown1`). Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name:

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "alien-mollusca-checklist", 
  "taxon",
  vdigest(scientific_name, algo = "md5"),
  sep = ":"
))
```

## Preview data

Show the number of taxa and distributions per rank:

```{r}
input_data %>%
  group_by(rankmarker) %>%
  summarize(
    `# taxa` = n_distinct(taxon_id),
    `# distributions` = n()
  ) %>%
  adorn_totals("row")
```

Preview data:

```{r}
input_data %>% head()
```

# Taxon core

## Pre-processing

Create a dataframe with unique taxa:

```{r}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

## Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records).

### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "Royal Belgian Institute of Natural Sciences")
```

### bibliographicCitation

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "")
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "RBINS")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "Registry of introduced alien molluscs in Belgium")
```

The following terms contain information about the taxon:

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name)
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = "Animalia")
```

### phylum

```{r}
taxon %<>% mutate(dwc_phylum = "Mollusca") 
```

### class

```{r}
taxon %<>% mutate(dwc_class = class)
```

### order

```{r}
taxon %<>% mutate(dwc_order = order) 
```

### family

```{r}
taxon %<>% mutate(dwc_family = family) 
```

### taxonRank

Inspect values:

```{r}
taxon %>%
  group_by(rankmarker) %>%
  count()
```

Map values by recoding to the [GBIF rank vocabulary](http://rs.gbif.org/vocabulary/gbif/rank_2015-04-24.xml):

```{r}
taxon %<>% mutate(dwc_taxonRank = recode(rankmarker,
  "sp."       = "species",
  .default    = "",
  .missing    = ""
))
```

Inspect mapped values: 

```{r}
taxon %>%
  group_by(rankmarker, dwc_taxonRank) %>%
  count()
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = "ICZN")
```

### taxonRemarks

## Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
```